---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path  = "man/figures/README-",
  fig.align = "center",
  out.width = "80%",
  dev       = "ragg_png",
  dpi       = 132
)
```

# ggplot tricks

<!-- badges: start -->
<!-- badges: end -->

The goal of this repository is to keep track of some neat [ggplot2](https://ggplot2.tidyverse.org/) tricks I've learned. This assumes you've familiarised yourself with the basics of ggplot2 and can construct some nice plots of your own. If not, please peruse the [book](https://ggplot2-book.org/) at your leasure.

I'm not incredibly adapt in gloriously typesetting plots and expertly finetuning themes and colour palettes, so you'd have to forgive me. The `mpg` dataset is very versatile for plotting, so you'll be seeing a lot of that as you read on. Extension packages are great, and I've dabbled myself, but I'll try to limit myself to vanilla ggplot2 tricks here.

For now, this will be mostly a README-only bag of tricks, but I may decide later to put them into separate groups in other files.

## Let's begin

By loading the library and setting a plotting theme. 
The first trick here is to use `theme_set()` to set a theme for *all* your plots throughout a document. If you find yourself setting a very verbose theme for every plot, here is the place where you set all your common settings. Then never write a novel of theme elements ever again!

```{r}
library(ggplot2)
library(scales)

theme_set(
  # Pick a starting theme
  theme_gray() +
  # Add your favourite elements
  theme(
    axis.line        = element_line(),
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line("grey95", linewidth = 0.25),
    legend.key       = element_rect(fill = NA) 
  )
)
```

## Half-geoms

There are some extensions that offer half-geom versions of things. Of the ones I know, [gghalves](https://erocoar.github.io/gghalves/) and the [see](https://easystats.github.io/see/) package offer some half-geoms.

Here is how to abuse the [delayed evaluation system](https://ggplot2.tidyverse.org/reference/aes_eval.html) to make your own. This can come in handy if you're not willing to take on an extra dependency for just this feature.

### Half-boxplots

The easy case is the boxplot. You can either set `xmin` or `xmax` to `after_scale(x)` to keep the right and left parts of a boxplot respectively. This still works fine with `position = "dodge"`.

```{r half_boxplot}
# A basic plot to reuse for examples
p <- ggplot(mpg, aes(class, displ, colour = class)) +
  guides(colour = "none", fill = "none") +
  labs(y = "Engine Displacement [L]", x = "Type of car")

p + geom_boxplot(aes(xmin = after_scale(x)))
```

### Half-errorbars

The same thing that works for boxplots, also works for errorbars.

```{r half_errorbar}
p + geom_errorbar(
  stat = "summary",
  fun.data = mean_se,
  aes(xmin = after_scale(x))
)
```

### Half-violin

We can once again do the same thing for violin plots, but the layer complains about not knowing about the `xmin` aesthetic. It does use that aesthetic, but only after the data has been setup, so it is not *intended* to be a user accessible aesthetic. We can silence the warning by updating the `xmin` default to `NULL`, which means it won't complain, but also doesn't use it if absent. 

```{r half_violin}
update_geom_defaults("violin", list(xmin = NULL))

p + geom_violin(aes(xmin = after_scale(x)))
```

### Combining

Not left as an exercise for the reader this time, but I just wanted to show how it would work if you were to combine two halves and want them a little bit offset from one another. We'll abuse the errorbars to serve as staples for the boxplots.

```{r combine_halves}
# A small nudge offset
offset <- 0.025

# We can pre-specify the mappings if we plan on recycling some
right_nudge <- aes(
  xmin = after_scale(x), 
  x = stage(class, after_stat = x + offset)
)
left_nudge  <- aes(
  xmax = after_scale(x),
  x = stage(class, after_stat = x - offset)
)

# Combining
p +
  geom_violin(right_nudge) +
  geom_boxplot(left_nudge) +
  geom_errorbar(left_nudge, stat = "boxplot", width = 0.3)
```

## Setting midpoints in divergent scales

Let's say you have better colour intuition than I have, and three colours aren't enough for your divergent colour palette needs. A painpoint is that it is tricky to get the midpoint right if your limits aren't perfectly centered around it. Enter the `rescaler` argument in league with `scales::rescale_mid()`.

```{r divergent_midpoint}
my_palette <- c("dodgerblue", "deepskyblue", "white", "hotpink", "deeppink")

ggplot(mpg, aes(displ, hwy, colour = cty - mean(cty))) +
  geom_point() +
  scale_colour_gradientn(
    colours = my_palette, 
    rescaler = ~ rescale_mid(.x, mid = 0)
  )
```


  <details style="margin-bottom:10px;">
  <summary>
  Session info
</summary>

```{r, echo = FALSE}
si <- sessioninfo::session_info()

# People don't need to see my file paths etc
si$platform$pandoc <- gsub(" @ C:/.*", "", si$platform$pandoc)
si$packages$library <- NULL

print(si)
```

</details>
